---
output:
  md_document:
    variant: gfm
---

**Course Description**

*Learn R Programming*

R programming language is a useful tool for data scientists, analysts, and statisticians, especially those working in academic settings. R's ability to handle complex analyses such as machine learning, financial modeling, and more makes it a valuable asset for a wide range of data-related tasks.

This introduction to R course covers the basics of this open source language, including vectors, factors, lists, and data frames. You’ll gain useful coding skills and be ready to start your own data analysis in R.

*Gain an Introduction to R*

You’ll get started with basic operations, like using the console as a calculator and understanding basic data types in R. Once you’ve had a chance to practice, you’ll move on to creating vectors and try out your new R skills on a data set based on betting in Las Vegas.

Next, you’ll learn how to work with matrices in R, learning how to create them, and perform calculations with them. You’ll also examine how R uses factors to store categorical data. Finally, you’ll explore how to work with R data frames and lists.

*Master the R Basics for Data Analysis*

By the time you’ve completed our Introduction to R course, you’ll be able to use R for your own data analysis. These sought-after skills can help you progress in your career and set you up for further learning. This course is part of several tracks, including Data Analyst with R, Data Scientist with R, and R Programming, all of which can help you develop your knowledge.

# Intro to basics

Take your first steps with R. In this chapter, you will learn how to use the console as a calculator and how to assign variables. You will also get to know the basic data types in R. Let's get started.

## How it works

In the editor on the right you should type R code to solve the
exercises. When you hit the 'Submit Answer' button, every line of code
is interpreted and executed by R and you get a message whether or not
your code was correct. The output of your R code is shown in the console
in the lower right corner.

R makes use of the `#` sign to add comments, so that you and others can
understand what the R code is about. Just like Twitter! Comments are not
run as R code, so they will not influence your result. For example,
*Calculate 3 + 4* in the editor on the right is a comment.

You can also execute R commands straight in the console. This is a good
way to experiment with R code, as your submission is not checked for
correctness.

-   In the editor on the right there is already some sample code. Can
    you see which lines are actual R code and which are comments?
-   Add a line of code that calculates the sum of 6 and 12, and hit the
    'Submit Answer' button.

```{r, eval=F}
# Calculate 3 + 4
3 + 4

# Calculate 6 + 12
6 + 12
```

## Arithmetic with R

In its most basic form, R can be used as a simple calculator. Consider
the following arithmetic operators:

-   Addition: `+`
-   Subtraction: `-`
-   Multiplication: `*`
-   Division: `/`
-   Exponentiation: `^`
-   Modulo: `%%`

The last two might need some explaining:

-   The `^` operator raises the number to its left to the power of the
    number to its right: for example `3^2` is 9.
-   The modulo returns the remainder of the division of the number to
    the left by the number on its right, for example 5 modulo 3 or
    `5 %% 3` is 2.

With this knowledge, follow the instructions to complete the exercise.

-   Type `2^5` in the editor to calculate 2 to the power 5.
-   Type `28 %% 6` to calculate 28 modulo 6.
-   Submit the answer and have a look at the R output in the console.
-   Note how the `#` symbol is used to add comments on the R code.

```{r, eval=F}
# An addition
5 + 5

# A subtraction
5 - 5 

# A multiplication
3 * 5

 # A division
(5 + 5) / 2 

# Exponentiation
2 ^ 5

# Modulo
28 %% 6
```

## Variable assignment

A basic concept in (statistical) programming is called a **variable**.

A variable allows you to store a value (e.g. 4) or an object (e.g. a
function description) in R. You can then later use this variable's name
to easily access the value or the object that is stored within this
variable.

You can assign a value 4 to a variable `my_var` with the command

    my_var <- 4

Over to you: complete the code in the editor such that it assigns the
value 42 to the variable `x` in the editor. Submit the answer. Notice
that when you ask R to print `x`, the value 42 appears.

```{r, eval=F}
# Assign the value 42 to x
x <- 42

# Print out the value of the variable x
x
```

## Variable assignment (2)

Suppose you have a fruit basket with five apples. As a data analyst in
training, you want to store the number of apples in a variable with the
name `my_apples`.

-   Type the following code in the editor: `my_apples <- 5`. This will
    assign the value 5 to `my_apples`.
-   Type: `my_apples` below the second comment. This will print out the
    value of `my_apples`.
-   Submit your answer, and look at the output: you see that the number
    5 is printed. So R now links the variable `my_apples` to the value
    5.

```{r, eval=F}
# Assign the value 5 to the variable my_apples
my_apples <- 5

# Print out the value of the variable my_apples
my_apples
```

## Variable assignment (3)

Every tasty fruit basket needs oranges, so you decide to add six
oranges. As a data analyst, your reflex is to immediately create the
variable `my_oranges` and assign the value 6 to it. Next, you want to
calculate how many pieces of fruit you have in total. Since you have
given meaningful names to these values, you can now code this in a clear
way:

    my_apples + my_oranges

-   Assign to `my_oranges` the value 6.
-   Add the variables `my_apples` and `my_oranges` and have R simply
    print the result.
-   Assign the result of adding `my_apples` and `my_oranges` to a new
    variable `my_fruit`.

```{r, eval=F}
# Assign a value to the variables my_apples and my_oranges
my_apples  <- 5
my_oranges <- 6

# Add these two variables together
my_apples + my_oranges

# Create the variable my_fruit
my_fruit <- my_apples + my_oranges
```

## Apples and oranges

Common knowledge tells you not to add apples and oranges. But hey, that
is what you just did, no :-)? The `my_apples` and `my_oranges` variables
both contained a number in the previous exercise. The `+` operator works
with numeric variables in R. If you really tried to add "apples" and
"oranges", and assigned a text value to the variable `my_oranges` (see
the editor), you would be trying to assign the addition of a numeric and
a character variable to the variable `my_fruit`. This is not possible.

-   Submit the answer and read the error message. Make sure to
    understand why this did not work.
-   Adjust the code so that R knows you have 6 oranges and thus a fruit
    basket with 11 pieces of fruit.

```{r, eval=F}
# Assign a value to the variable my_apples
my_apples <- 5  

# Fix the assignment of my_oranges
my_oranges <- 6

# Create the variable my_fruit and print it out
my_fruit <- my_apples + my_oranges 
my_fruit
```

## Basic data types in R

R works with numerous data types. Some of the most basic types to get
started are:

-   Decimal values like `4.5` are called **numerics**.
-   Whole numbers like `4` are called **integers**. Integers are also
    numerics.
-   Boolean values (`TRUE` or `FALSE`) are called **logical**.
-   Text (or string) values are called **characters**.

Note how the quotation marks in the editor indicate that `"some text"`
is a string.

Change the value of the:

-   `my_numeric` variable to `42`.
-   `my_character` variable to `"universe"`. Note that the quotation
    marks indicate that `"universe"` is a character.
-   `my_logical` variable to `FALSE`.

Note that R is case sensitive!

```{r, eval=F}
# Change my_numeric to be 42
my_numeric <- 42

# Change my_character to be "universe"
my_character <- "universe"

# Change my_logical to be FALSE
my_logical <- FALSE
```

## What's that data type?

Do you remember that when you added `5 + "six"`, you got an error due to
a mismatch in data types? You can avoid such embarrassing situations by
checking the data type of a variable beforehand. You can do this with
the `class()` function, as the code in the editor shows.

Complete the code in the editor and also print out the classes of
`my_character` and `my_logical`.

```{r, eval=F}
# Declare variables of different types:
my_numeric <- 42
my_character <- "universe"
my_logical <- FALSE

# Check class of my_numeric
class(my_numeric)

# Check class of my_character
class(my_character)

# Check class of my_logical
class(my_logical)
```

# Vectors

We take you on a trip to Vegas, where you will learn how to analyze your gambling results using vectors in R. After completing this chapter, you will be able to create vectors in R, name them, select elements from them, and compare different vectors.

## Create a vector

Feeling lucky? You better, because this chapter takes you on a trip to
the City of Sins, also known as *Statisticians Paradise*!

Thanks to R and your new data-analytical skills, you will learn how to
uplift your performance at the tables and fire off your career as a
professional gambler. This chapter will show how you can easily keep
track of your betting progress and how you can do some simple analyses
on past actions. Next stop, Vegas Baby… VEGAS!!

-   Do you still remember what you have learned in the first chapter?
    Assign the value `"Go!"` to the variable `vegas`. Remember: R is
    case sensitive!

```{r, eval=F}
# Define the variable vegas
vegas <- "Go!"
```

## Create a vector (2)

Let us focus first!

On your way from rags to riches, you will make extensive use of vectors.
Vectors are one-dimension arrays that can hold numeric data, character
data, or logical data. In other words, a vector is a simple tool to
store data. For example, you can store your daily gains and losses in
the casinos.

In R, you create a vector with the combine function
[`c()`](http://www.rdocumentation.org/packages/base/functions/c). You
place the vector elements separated by a comma between the parentheses.
For example:

    numeric_vector <- c(1, 2, 3)
    character_vector <- c("a", "b", "c")

Once you have created these vectors in R, you can use them to do
calculations.

Complete the code such that `boolean_vector` contains the three
elements: `TRUE`, `FALSE` and `TRUE` (in that order).

```{r, eval=F}
numeric_vector <- c(1, 10, 49)
character_vector <- c("a", "b", "c")

# Complete the code for boolean_vector
boolean_vector <- c(TRUE, FALSE, TRUE)
```

## Create a vector (3)

After one week in Las Vegas and still zero Ferraris in your garage, you
decide that it is time to start using your data analytical superpowers.

Before doing a first analysis, you decide to first collect all the
winnings and losses for the last week:

For `poker_vector`:

-   On Monday you won $140
-   Tuesday you lost $50
-   Wednesday you won $20
-   Thursday you lost $120
-   Friday you won $240

For `roulette_vector`:

-   On Monday you lost $24
-   Tuesday you lost $50
-   Wednesday you won $100
-   Thursday you lost $350
-   Friday you won $10

You only played poker and roulette, since there was a delegation of
mediums that occupied the craps tables. To be able to use this data in
R, you decide to create the variables `poker_vector` and
`roulette_vector`.

Assign the winnings/losses for roulette to the variable
`roulette_vector`. You lost $24, then lost $50, won $100, lost $350, and
won $10.

```{r, eval=F}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)
```

## Naming a vector

As a data analyst, it is important to have a clear view on the data that
you are using. Understanding what each element refers to is therefore
essential.

In the previous exercise, we created a vector with your winnings over
the week. Each vector element refers to a day of the week but it is hard
to tell which element belongs to which day. It would be nice if you
could show that in the vector itself.

You can give a name to the elements of a vector with the `names()`
function. Have a look at this example:

    some_vector <- c("John Doe", "poker player")
    names(some_vector) <- c("Name", "Profession")

This code first creates a vector `some_vector` and then gives the two
elements a name. The first element is assigned the name `Name`, while
the second element is labeled `Profession`. Printing the contents to the
console yields following output:

              Name     Profession 
        "John Doe" "poker player" 

-   The code in the editor names the elements in `poker_vector` with the
    days of the week. Add code to do the same thing for
    `roulette_vector`.

```{r, eval=F}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Assign days as names of poker_vector
names(poker_vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

# Assign days as names of roulette_vector
names(roulette_vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
```

## Naming a vector (2)

If you want to become a good statistician, you have to become lazy. (If
you are already lazy, chances are high you are one of those exceptional,
natural-born statistical talents.)

In the previous exercises you probably experienced that it is boring and
frustrating to type and retype information such as the days of the week.
However, when you look at it from a higher perspective, there is a more
efficient way to do this, namely, to assign the days of the week vector
to a **variable**!

Just like you did with your poker and roulette returns, you can also
create a variable that contains the days of the week. This way you can
use and re-use it.

-   A variable `days_vector` that contains the days of the week has
    already been created for you.
-   Use `days_vector` to set the names of `poker_vector` and
    `roulette_vector`.

```{r, eval=F}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# The variable days_vector
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

# Assign the names of the day to roulette_vector and poker_vector
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector
```

## Calculating total winnings

Now that you have the poker and roulette winnings nicely as named
vectors, you can start doing some data analytical magic.

You want to find out the following type of information:

-   How much has been your overall profit or loss per day of the week?
-   Have you lost money over the week in total?
-   Are you winning/losing money on poker or on roulette?

To get the answers, you have to do arithmetic calculations on vectors.

It is important to know that if you sum two vectors in R, it takes the
element-wise sum. For example, the following three statements are
completely equivalent:

    c(1, 2, 3) + c(4, 5, 6)
    c(1 + 4, 2 + 5, 3 + 6)
    c(5, 7, 9)

You can also do the calculations with variables that represent vectors:

    a <- c(1, 2, 3) 
    b <- c(4, 5, 6)
    c <- a + b

-   Take the sum of the variables `A_vector` and `B_vector` and assign
    it to `total_vector`.
-   Inspect the result by printing out `total_vector`.

```{r, eval=F}
A_vector <- c(1, 2, 3)
B_vector <- c(4, 5, 6)

# Take the sum of A_vector and B_vector
total_vector <- A_vector + B_vector

# Print out total_vector
total_vector
```

## Calculating total winnings (2)

Now you understand how R does arithmetic with vectors, it is time to get
those Ferraris in your garage! First, you need to understand what the
overall profit or loss per day of the week was. The total daily profit
is the sum of the profit/loss you realized on poker per day, and the
profit/loss you realized on roulette per day.

In R, this is just the sum of `roulette_vector` and `poker_vector`.

Assign to the variable `total_daily` how much you won or lost on each
day in total (poker and roulette combined).

```{r, eval=F}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Assign to total_daily how much you won/lost on each day
total_daily <- poker_vector + roulette_vector
```

## Calculating total winnings (3)

Based on the previous analysis, it looks like you had a mix of good and
bad days. This is not what your ego expected, and you wonder if there
may be a very tiny chance you have lost money over the week in total?

A function that helps you to answer this question is
[`sum()`](http://www.rdocumentation.org/packages/base/functions/sum). It
calculates the sum of all elements of a vector. For example, to
calculate the total amount of money you have lost/won with poker you do:

    total_poker <- sum(poker_vector)

-   Calculate the total amount of money that you have won/lost with
    roulette and assign to the variable `total_roulette`.
-   Now that you have the totals for roulette and poker, you can easily
    calculate `total_week` (which is the sum of all gains and losses of
    the week).
-   Print out `total_week`.

```{r, eval=F}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Total winnings with poker
total_poker <- sum(poker_vector)

# Total winnings with roulette
total_roulette <-  sum(roulette_vector)

# Total winnings overall
total_week <- total_roulette + total_poker

# Print out total_week
total_week
```

## Comparing total winnings

Oops, it seems like you are losing money. Time to rethink and adapt your
strategy! This will require some deeper analysis…

After a short brainstorm in your hotel's jacuzzi, you realize that a
possible explanation might be that your skills in roulette are not as
well developed as your skills in poker. So maybe your total gains in
poker are higher (or `>` ) than in roulette.

-   Calculate `total_poker` and `total_roulette` as in the previous
    exercise. Use the `sum()` function twice.
-   Check if your total gains in poker are higher than for roulette by
    using a comparison. Simply print out the result of this comparison.
    What do you conclude, should you focus on roulette or on poker?

```{r, eval=F}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Calculate total gains for poker and roulette
total_poker <- sum(poker_vector)
total_roulette <- sum(roulette_vector)

# Check if you realized higher total gains in poker than in roulette
total_poker > total_roulette
```

## Vector selection: the good times

Your hunch seemed to be right. It appears that the poker game is more
your cup of tea than roulette.

Another possible route for investigation is your performance at the
beginning of the working week compared to the end of it. You did have a
couple of Margarita cocktails at the end of the week…

To answer that question, you only want to focus on a selection of the
`total_vector`. In other words, our goal is to select specific elements
of the vector. To select elements of a vector (and later matrices, data
frames, …), you can use square brackets. Between the square brackets,
you indicate what elements to select. For example, to select the first
element of the vector, you type `poker_vector[1]`. To select the second
element of the vector, you type `poker_vector[2]`, etc. Notice that the
first element in a vector has index 1, not 0 as in many other
programming languages.

Assign the poker results of Wednesday to the variable `poker_wednesday`.

```{r, eval=F}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Define a new variable based on a selection
poker_wednesday <- poker_vector[3]
```

## Vector selection: the good times (2)

How about analyzing your midweek results?

To select multiple elements from a vector, you can add square brackets
at the end of it. You can indicate between the brackets what elements
should be selected. For example: suppose you want to select the first
and the fifth day of the week: use the vector `c(1, 5)` between the
square brackets. For example, the code below selects the first and fifth
element of `poker_vector`:

    poker_vector[c(1, 5)]

Assign the poker results of Tuesday, Wednesday and Thursday to the
variable `poker_midweek`.

```{r, eval=F}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Define a new variable based on a selection
poker_midweek <- poker_vector[c(2, 3, 4)]
```

## Vector selection: the good times (3)

Selecting multiple elements of `poker_vector` with `c(2, 3, 4)` is not
very convenient. Many statisticians are lazy people by nature, so they
created an easier way to do this: `c(2, 3, 4)` can be abbreviated
to`2:4`, which generates a vector with all natural numbers from 2 up to
4.

So, another way to find the mid-week results is `poker_vector[2:4]`.
Notice how the vector `2:4` is placed between the square brackets to
select element 2 up to 4.

Assign to `roulette_selection_vector` the roulette results from Tuesday
up to Friday; make use of `:` if it makes things easier for you.

```{r, eval=F}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Define a new variable based on a selection
roulette_selection_vector <- roulette_vector[2:5]
```

## Vector selection: the good times (4)

Another way to tackle the previous exercise is by using the names of the
vector elements (Monday, Tuesday, …) instead of their numeric positions.
For example,

    poker_vector["Monday"]

will select the first element of `poker_vector` since `"Monday"` is the
name of that first element.

Just like you did in the previous exercise with numerics, you can also
use the element names to select multiple elements, for example:

    poker_vector[c("Monday","Tuesday")]

-   Select the first three elements in `poker_vector` by using their
    names: `"Monday"`, `"Tuesday"` and `"Wednesday"`. Assign the result
    of the selection to `poker_start`.
-   Calculate the average of the values in `poker_start` with the
    [`mean()`](http://www.rdocumentation.org/packages/base/functions/mean)
    function. Simply print out the result so you can inspect it.

```{r, eval=F}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Select poker results for Monday, Tuesday and Wednesday
poker_start <- poker_vector[c("Monday", "Tuesday", "Wednesday")]
  
# Calculate the average of the elements in poker_start
mean(poker_start)
```

## Selection by comparison - Step 1

By making use of comparison operators, we can approach the previous
question in a more proactive way.

The (logical) comparison operators known to R are:

-   `<` for less than
-   `>` for greater than
-   `<=` for less than or equal to
-   `>=` for greater than or equal to
-   `==` for equal to each other
-   `!=` not equal to each other

As seen in the previous chapter, stating `6 > 5` returns `TRUE`. The
nice thing about R is that you can use these comparison operators also
on vectors. For example:

    c(4, 5, 6) > 5
    [1] FALSE FALSE TRUE

This command tests for every element of the vector if the condition
stated by the comparison operator is `TRUE` or `FALSE`.

-   Check which elements in `poker_vector` are positive (i.e. \> 0) and
    assign this to `selection_vector`.
-   Print out `selection_vector` so you can inspect it. The printout
    tells you whether you won (`TRUE`) or lost (`FALSE`) any money for
    each day.

```{r, eval=F}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on poker?
selection_vector <- poker_vector > 0
  
# Print out selection_vector
selection_vector
```

## Selection by comparison - Step 2

Working with comparisons will make your data analytical life easier.
Instead of selecting a subset of days to investigate yourself (like
before), you can simply ask R to return only those days where you
realized a positive return for poker.

In the previous exercises you used
`selection_vector <- poker_vector > 0` to find the days on which you had
a positive poker return. Now, you would like to know not only the days
on which you won, but also how much you won on those days.

You can select the desired elements, by putting `selection_vector`
between the square brackets that follow `poker_vector`:

    poker_vector[selection_vector]

R knows what to do when you pass a logical vector in square brackets: it
will only select the elements that correspond to `TRUE` in
`selection_vector`.

Use `selection_vector` in square brackets to assign the amounts that you
won on the profitable days to the variable `poker_winning_days`.

```{r, eval=F}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on poker?
selection_vector <- poker_vector > 0

# Select from poker_vector these days
poker_winning_days <- poker_vector[selection_vector]
```

## Advanced selection

Just like you did for poker, you also want to know those days where you
realized a positive return for roulette.

-   Create the variable `selection_vector`, this time to see if you made
    profit with roulette for different days.
-   Assign the amounts that you made on the days that you ended
    positively for roulette to the variable `roulette_winning_days`.
    This vector thus contains the positive winnings of
    `roulette_vector`.

```{r, eval=F}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on roulette?
selection_vector <- roulette_vector > 0

# Select from roulette_vector these days
roulette_winning_days <- roulette_vector[selection_vector]
```

# Matrices

In this chapter, you will learn how to work with matrices in R. By the end of the chapter, you will be able to create matrices and understand how to do basic computations with them. You will analyze the box office numbers of the Star Wars movies and learn how to use matrices in R. May the force be with you!

## What's a matrix?

In R, a matrix is a collection of elements of the same data type
(numeric, character, or logical) arranged into a fixed number of rows
and columns. Since you are only working with rows and columns, a matrix
is called two-dimensional.

You can construct a matrix in R with the
[`matrix()`](http://www.rdocumentation.org/packages/base/functions/matrix)
function. Consider the following example:

    matrix(1:9, byrow = TRUE, nrow = 3)

In the
[`matrix()`](http://www.rdocumentation.org/packages/base/functions/matrix)
function:

-   The first argument is the collection of elements that R will arrange
    into the rows and columns of the matrix. Here, we use `1:9` which is
    a shortcut for `c(1, 2, 3, 4, 5, 6, 7, 8, 9)`.
-   The argument `byrow` indicates that the matrix is filled by the
    rows. If we want the matrix to be filled by the columns, we just
    place `byrow = FALSE`.
-   The third argument `nrow` indicates that the matrix should have
    three rows.

Construct a matrix with 3 rows containing the numbers 1 up to 9, filled
row-wise.

```{r, eval=F}
# Construct a matrix with 3 rows that contain the numbers 1 up to 9
matrix(1:9, byrow = TRUE, nrow = 3)
```

## Analyze matrices, you shall

It is now time to get your hands dirty. In the following exercises you
will analyze the box office numbers of the Star Wars franchise. May the
force be with you!

In the editor, three vectors are defined. Each one represents the box
office numbers from the first three Star Wars movies. The first element
of each vector indicates the US box office revenue, the second element
refers to the Non-US box office (source: Wikipedia).

In this exercise, you'll combine all these figures into a single vector.
Next, you'll build a matrix from this vector.

-   Use `c(new_hope, empire_strikes, return_jedi)` to combine the three
    vectors into one vector. Call this vector `box_office`.
-   Construct a matrix with 3 rows, where each row represents a movie.
    Use the `matrix()` function to do this. The first argument is the
    vector `box_office`, containing all box office figures. Next, you'll
    have to specify `nrow = 3` and `byrow = TRUE`. Name the resulting
    matrix `star_wars_matrix`.

```{r, eval=F}
# Box office Star Wars (in millions!)
new_hope <- c(460.998, 314.4)
empire_strikes <- c(290.475, 247.900)
return_jedi <- c(309.306, 165.8)

# Create box_office
box_office <- c(new_hope, empire_strikes, return_jedi)

# Construct star_wars_matrix
star_wars_matrix <- matrix(box_office, nrow = 3, byrow = TRUE) 
```

## Naming a matrix

To help you remember what is stored in `star_wars_matrix`, you would
like to add the names of the movies for the rows. Not only does this
help you to read the data, but it is also useful to select certain
elements from the matrix.

Similar to vectors, you can add names for the rows and the columns of a
matrix

    rownames(my_matrix) <- row_names_vector
    colnames(my_matrix) <- col_names_vector

We went ahead and prepared two vectors for you: `region`, and `titles`.
You will need these vectors to name the columns and rows of
`star_wars_matrix`, respectively.

-   Use `colnames()` to name the columns of `star_wars_matrix` with the
    `region` vector.
-   Use `rownames()` to name the rows of `star_wars_matrix` with the
    `titles` vector.
-   Print out `star_wars_matrix` to see the result of your work.

```{r, eval=F}
# Box office Star Wars (in millions!)
new_hope <- c(460.998, 314.4)
empire_strikes <- c(290.475, 247.900)
return_jedi <- c(309.306, 165.8)

# Construct matrix
star_wars_matrix <- matrix(c(new_hope, empire_strikes, return_jedi), nrow = 3, byrow = TRUE)

# Vectors region and titles, used for naming
region <- c("US", "non-US")
titles <- c("A New Hope", "The Empire Strikes Back", "Return of the Jedi")

# Name the columns with region
colnames(star_wars_matrix) <- region

# Name the rows with titles
rownames(star_wars_matrix) <- titles

# Print out star_wars_matrix
star_wars_matrix
```

## Calculating the worldwide box office

The single most important thing for a movie in order to become an
instant legend in Tinseltown is its worldwide box office figures.

To calculate the total box office revenue for the three Star Wars
movies, you have to take the sum of the US revenue column and the non-US
revenue column.

In R, the function
[`rowSums()`](http://www.rdocumentation.org/packages/base/functions/colSums)
conveniently calculates the totals for each row of a matrix. This
function creates a new vector:

    rowSums(my_matrix)

Calculate the worldwide box office figures for the three movies and put
these in the vector named `worldwide_vector`.

```{r, eval=F}
# Construct star_wars_matrix
box_office <- c(460.998, 314.4, 290.475, 247.900, 309.306, 165.8)
region <- c("US", "non-US")
titles <- c("A New Hope", 
            "The Empire Strikes Back", 
            "Return of the Jedi")
               
star_wars_matrix <- matrix(box_office, 
                      nrow = 3, byrow = TRUE,
                      dimnames = list(titles, region))

# Calculate worldwide box office figures
worldwide_vector <- rowSums(star_wars_matrix)
```

## Adding a column for the Worldwide box office

In the previous exercise you calculated the vector that contained the
worldwide box office receipt for each of the three Star Wars movies.
However, this vector is not yet part of `star_wars_matrix`.

You can add a column or multiple columns to a matrix with the
[`cbind()`](http://www.rdocumentation.org/packages/base/functions/cbind)
function, which merges matrices and/or vectors together by column. For
example:

    big_matrix <- cbind(matrix1, matrix2, vector1 ...)

Add `worldwide_vector` as a new column to the `star_wars_matrix` and
assign the result to `all_wars_matrix`. Use the
[`cbind()`](http://www.rdocumentation.org/packages/base/functions/cbind)
function.

```{r, eval=F}
# Construct star_wars_matrix
box_office <- c(460.998, 314.4, 290.475, 247.900, 309.306, 165.8)
region <- c("US", "non-US")
titles <- c("A New Hope", 
            "The Empire Strikes Back", 
            "Return of the Jedi")
               
star_wars_matrix <- matrix(box_office, 
                      nrow = 3, byrow = TRUE,
                      dimnames = list(titles, region))

# The worldwide box office figures
worldwide_vector <- rowSums(star_wars_matrix)

# Bind the new variable worldwide_vector as a column to star_wars_matrix
all_wars_matrix <- cbind(star_wars_matrix, worldwide_vector)
```

## Adding a row

Just like every action has a reaction, every
[`cbind()`](http://www.rdocumentation.org/packages/base/functions/cbind)
has an
[`rbind()`](http://www.rdocumentation.org/packages/base/functions/cbind).
(We admit, we are pretty bad with metaphors.)

Your R workspace, where all variables you defined 'live' ([check out
what a workspace
is](http://www.statmethods.net/interface/workspace.html)), has already
been initialized and contains two matrices:

-   `star_wars_matrix` that we have used all along, with data on the
    original trilogy,
-   `star_wars_matrix2`, with similar data for the prequels trilogy.

Explore these matrices in the console if you want to have a closer look.
If you want to check out the contents of the workspace, you can type
`ls()` in the console.

Use `rbind()` to paste together `star_wars_matrix` and
`star_wars_matrix2`, in this order. Assign the resulting matrix to
`all_wars_matrix`.

```{r, eval=F}
# edited/added
box_office2 <- c(474.5, 552.5, 310.7, 338.7, 380.3, 468.5)
region2 <- c("US", "non-US")
titles2 <- c("The Phantom Menace", 
            "Attack of the Clones", 
            "Revenge of the Sith")
star_wars_matrix2 <- matrix(box_office2, 
                      nrow = 3, byrow = TRUE,
                      dimnames = list(titles2, region2))

# star_wars_matrix and star_wars_matrix2 are available in your workspace
star_wars_matrix  
star_wars_matrix2 

# Combine both Star Wars trilogies in one matrix
all_wars_matrix <- rbind(star_wars_matrix, star_wars_matrix2)
```

## The total box office revenue for the entire saga

Just like
[`cbind()`](http://www.rdocumentation.org/packages/base/functions/cbind)
has
[`rbind()`](http://www.rdocumentation.org/packages/base/functions/cbind),
[`colSums()`](http://www.rdocumentation.org/packages/base/functions/colSums)
has
[`rowSums()`](http://www.rdocumentation.org/packages/base/functions/colSums).
Your R workspace already contains the `all_wars_matrix` that you
constructed in the previous exercise; type `all_wars_matrix` to have
another look. Let's now calculate the total box office revenue for the
entire saga.

-   Calculate the total revenue for the US and the non-US region and
    assign `total_revenue_vector`. You can use the
    [`colSums()`](http://www.rdocumentation.org/packages/base/functions/colSums)
    function.
-   Print out `total_revenue_vector` to have a look at the results.

```{r, eval=F}
# all_wars_matrix is available in your workspace
all_wars_matrix

# Total revenue for US and non-US
total_revenue_vector <- colSums(all_wars_matrix)

# Print out total_revenue_vector
total_revenue_vector
```

## Selection of matrix elements

Similar to vectors, you can use the square brackets `[ ]` to select one
or multiple elements from a matrix. Whereas vectors have one dimension,
matrices have two dimensions. You should therefore use a comma to
separate the rows you want to select from the columns. For example:

-   `my_matrix[1,2]` selects the element at the first row and second
    column.
-   `my_matrix[1:3,2:4]` results in a matrix with the data on the rows
    1, 2, 3 and columns 2, 3, 4.

If you want to select all elements of a row or a column, no number is
needed before or after the comma, respectively:

-   `my_matrix[,1]` selects all elements of the first column.
-   `my_matrix[1,]` selects all elements of the first row.

Back to Star Wars with this newly acquired knowledge! As in the previous
exercise, `all_wars_matrix` is already available in your workspace.

-   Select the non-US revenue for all movies (the entire second column
    of `all_wars_matrix`), store the result as `non_us_all`.
-   Use `mean()` on `non_us_all` to calculate the average non-US revenue
    for all movies. Simply print out the result.
-   This time, select the non-US revenue for the first two movies in
    `all_wars_matrix`. Store the result as `non_us_some`.
-   Use `mean()` again to print out the average of the values in
    `non_us_some`.

```{r, eval=F}
# all_wars_matrix is available in your workspace
all_wars_matrix

# Select the non-US revenue for all movies
non_us_all <- all_wars_matrix[,2]
  
# Average non-US revenue
mean(non_us_all)
  
# Select the non-US revenue for first two movies
non_us_some <- all_wars_matrix[1:2,2]
  
# Average non-US revenue for first two movies
mean(non_us_some)
```

## A little arithmetic with matrices

Similar to what you have learned with vectors, the standard operators
like `+`, `-`, `/`, `*`, etc. work in an element-wise way on matrices in
R.

For example, `2 * my_matrix` multiplies each element of `my_matrix` by
two.

As a newly-hired data analyst for Lucasfilm, it is your job to find out
how many visitors went to each movie for each geographical area. You
already have the total revenue figures in `all_wars_matrix`. Assume that
the price of a ticket was 5 dollars. Simply dividing the box office
numbers by this ticket price gives you the number of visitors.

-   Divide `all_wars_matrix` by 5, giving you the number of visitors in
    millions. Assign the resulting matrix to `visitors`.
-   Print out `visitors` so you can have a look.

```{r, eval=F}
# all_wars_matrix is available in your workspace
all_wars_matrix

# Estimate the visitors
visitors <- all_wars_matrix / 5

# Print the estimate to the console
visitors
```

## A little arithmetic with matrices (2)

Just like `2 * my_matrix` multiplied every element of `my_matrix` by
two, `my_matrix1 * my_matrix2` creates a matrix where each element is
the product of the corresponding elements in `my_matrix1` and
`my_matrix2`.

After looking at the result of the previous exercise, big boss Lucas
points out that the ticket prices went up over time. He asks to redo the
analysis based on the prices you can find in `ticket_prices_matrix`
(source: imagination).

*Those who are familiar with matrices should note that this is not the
standard matrix multiplication for which you should use `%*%` in R.*

-   Divide `all_wars_matrix` by `ticket_prices_matrix` to get the
    estimated number of US and non-US visitors for the six movies.
    Assign the result to `visitors`.
-   From the `visitors` matrix, select the entire first column,
    representing the number of visitors in the US. Store this selection
    as `us_visitors`.
-   Calculate the average number of US visitors; print out the result.

```{r, eval=F}
# edited/added
ticket_prices1 <- c(5, 5, 6, 6, 7, 7)
region1 <- c("US", "non-US")
titles1 <- c("The Phantom Menace", 
            "Attack of the Clones", 
            "Revenge of the Sith")
region2 <- c("US", "non-US")
titles2 <- c("The Phantom Menace", 
            "Attack of the Clones", 
            "Revenge of the Sith")
ticket_prices2 <- c(4, 4, 4.5, 4.5, 4.9, 4.9)
ticket_prices_matrix1 <- matrix(ticket_prices1, 
                      nrow = 3, byrow = TRUE,
                      dimnames = list(titles1, region1))
ticket_prices_matrix2 <- matrix(ticket_prices2, 
                      nrow = 3, byrow = TRUE,
                      dimnames = list(titles2, region2))
ticket_prices_matrix <- rbind(ticket_prices_matrix1, ticket_prices_matrix2)

# all_wars_matrix and ticket_prices_matrix are available in your workspace
all_wars_matrix
ticket_prices_matrix

# Estimated number of visitors
visitors <- all_wars_matrix / ticket_prices_matrix

# US visitors
us_visitors <- visitors[ ,1]

# Average number of US visitors
mean(us_visitors)
```

# Factors

Data often falls into a limited number of categories. For example, human hair color can be categorized as black, brown, blond, red, grey, or white—and perhaps a few more options for people who color their hair. In R, categorical data is stored in factors. Factors are very important in data analysis, so start learning how to create, subset, and compare them now.

## What's a factor and why would you use it?

In this chapter you dive into the wonderful world of **factors**.

The term factor refers to a statistical data type used to store
categorical variables. The difference between a categorical variable and
a continuous variable is that a categorical variable can belong to a
**limited number of categories**. A continuous variable, on the other
hand, can correspond to an infinite number of values.

It is important that R knows whether it is dealing with a continuous or
a categorical variable, as the statistical models you will develop in
the future treat both types differently. (You will see later why this is
the case.)

A good example of a categorical variable is sex. In many circumstances
you can limit the sex categories to "Male" or "Female". (Sometimes you
may need different categories. For example, you may need to consider
chromosomal variation, hermaphroditic animals, or different cultural
norms, but you will always have a finite number of categories.)

Assign to variable `theory` the value `"factors"`.

```{r, eval=F}
# Assign to the variable theory what this chapter is about!
theory <- "factors"
```

## What's a factor and why would you use it? (2)

To create factors in R, you make use of the function
[`factor()`](http://www.rdocumentation.org/packages/base/functions/factor).
First thing that you have to do is create a vector that contains all the
observations that belong to a limited number of categories. For example,
`sex_vector` contains the sex of 5 different individuals:

    sex_vector <- c("Male","Female","Female","Male","Male")

It is clear that there are two categories, or in R-terms **'factor
levels'**, at work here: "Male" and "Female".

The function
[`factor()`](http://www.rdocumentation.org/packages/base/functions/factor)
will encode the vector as a factor:

    factor_sex_vector <- factor(sex_vector)

-   Convert the character vector `sex_vector` to a factor with
    `factor()` and assign the result to `factor_sex_vector`
-   Print out `factor_sex_vector` and assert that R prints out the
    factor levels below the actual values.

```{r, eval=F}
# Sex vector
sex_vector <- c("Male", "Female", "Female", "Male", "Male")

# Convert sex_vector to a factor
factor_sex_vector <- factor(sex_vector)

# Print out factor_sex_vector
factor_sex_vector
```

## What's a factor and why would you use it? (3)

There are two types of categorical variables: a **nominal categorical
variable** and an **ordinal categorical variable**.

A nominal variable is a categorical variable without an implied order.
This means that it is impossible to say that 'one is worth more than the
other'. For example, think of the categorical variable `animals_vector`
with the categories `"Elephant"`, `"Giraffe"`, `"Donkey"` and `"Horse"`.
Here, it is impossible to say that one stands above or below the other.
(Note that some of you might disagree ;-) ).

In contrast, ordinal variables do have a natural ordering. Consider for
example the categorical variable `temperature_vector` with the
categories: `"Low"`, `"Medium"` and `"High"`. Here it is obvious that
`"Medium"` stands above `"Low"`, and `"High"` stands above `"Medium"`.

Submit the answer to check how R constructs and prints nominal and
ordinal variables. Do not worry if you do not understand all the code
just yet, we will get to that.

```{r, eval=F}
# Animals
animals_vector <- c("Elephant", "Giraffe", "Donkey", "Horse")
factor_animals_vector <- factor(animals_vector)
factor_animals_vector

# Temperature
temperature_vector <- c("High", "Low", "High","Low", "Medium")
factor_temperature_vector <- factor(temperature_vector, order = TRUE, levels = c("Low", "Medium", "High"))
factor_temperature_vector
```

## Factor levels

When you first get a dataset, you will often notice that it contains
factors with specific factor levels. However, sometimes you will want to
change the names of these levels for clarity or other reasons. R allows
you to do this with the function
[`levels()`](http://www.rdocumentation.org/packages/base/functions/levels):

    levels(factor_vector) <- c("name1", "name2",...)

A good illustration is the raw data that is provided to you by a survey.
A common question for every questionnaire is the sex of the respondent.
Here, for simplicity, just two categories were recorded, `"M"` and
`"F"`. (You usually need more categories for survey data; either way,
you use a factor to store the categorical data.)

    survey_vector <- c("M", "F", "F", "M", "M")

Recording the sex with the abbreviations `"M"` and `"F"` can be
convenient if you are collecting data with pen and paper, but it can
introduce confusion when analyzing the data. At that point, you will
often want to change the factor levels to `"Male"` and `"Female"`
instead of `"M"` and `"F"` for clarity.

**Watch out:** the order with which you assign the levels is important.
If you type `levels(factor_survey_vector)`, you'll see that it outputs
`[1] "F" "M"`. If you don't specify the levels of the factor when
creating the vector, `R` will automatically assign them alphabetically.
To correctly map `"F"` to `"Female"` and `"M"` to `"Male"`, the levels
should be set to `c("Female", "Male")`, in this order.

-   Check out the code that builds a factor vector from `survey_vector`.
    You should use `factor_survey_vector` in the next instruction.
-   Change the factor levels of `factor_survey_vector` to
    `c("Female", "Male")`. Mind the order of the vector elements here.

```{r, eval=F}
# Code to build factor_survey_vector
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)

# Specify the levels of factor_survey_vector
levels(factor_survey_vector) <- c("Female", "Male")

factor_survey_vector
```

## Summarizing a factor

After finishing this course, one of your favorite functions in R will be
[`summary()`](http://www.rdocumentation.org/packages/base/functions/summary).
This will give you a quick overview of the contents of a variable:

    summary(my_var)

Going back to our survey, you would like to know how many `"Male"`
responses you have in your study, and how many `"Female"` responses. The
[`summary()`](http://www.rdocumentation.org/packages/base/functions/summary)
function gives you the answer to this question.

Ask a
[`summary()`](http://www.rdocumentation.org/packages/base/functions/summary)
of the `survey_vector` and `factor_survey_vector`. Interpret the results
of both vectors. Are they both equally useful in this case?

```{r, eval=F}
# Build factor_survey_vector with clean levels
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)
levels(factor_survey_vector) <- c("Female", "Male")
factor_survey_vector

# Generate summary for survey_vector
summary(survey_vector)

# Generate summary for factor_survey_vector
summary(factor_survey_vector)
```

## Battle of the sexes

You might wonder what happens when you try to compare elements of a
factor. In `factor_survey_vector` you have a factor with two levels:
`"Male"` and `"Female"`. But how does R value these relative to each
other?

Read the code in the editor and submit the answer to test if `male` is
greater than (`>`) `female`.

```{r, eval=F}
# Build factor_survey_vector with clean levels
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)
levels(factor_survey_vector) <- c("Female", "Male")

# Male
male <- factor_survey_vector[1]

# Female
female <- factor_survey_vector[2]

# Battle of the sexes: Male 'larger' than female?
male > female
```

## Ordered factors

Since `"Male"` and `"Female"` are unordered (or nominal) factor levels,
R returns a warning message, telling you that the greater than operator
is not meaningful. As seen before, R attaches an equal value to the
levels for such factors.

But this is not always the case! Sometimes you will also deal with
factors that do have a natural ordering between its categories. If this
is the case, we have to make sure that we pass this information to R…

Let us say that you are leading a research team of five data analysts
and that you want to evaluate their performance. To do this, you track
their speed, evaluate each analyst as `"slow"`, `"medium"` or `"fast"`,
and save the results in `speed_vector`.

As a first step, assign `speed_vector` a vector with 5 entries, one for
each analyst. Each entry should be either `"slow"`, `"medium"`, or
`"fast"`. Use the list below:

-   Analyst 1 is medium,
-   Analyst 2 is slow,
-   Analyst 3 is slow,
-   Analyst 4 is medium and
-   Analyst 5 is fast.

No need to specify these are factors yet.

```{r, eval=F}
# Create speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")
```

## Ordered factors (2)

`speed_vector` should be converted to an ordinal factor since its
categories have a natural ordering. By default, the function
[`factor()`](http://www.rdocumentation.org/packages/base/functions/factor)
transforms `speed_vector` into an unordered factor. To create an ordered
factor, you have to add two additional arguments: `ordered` and
`levels`.

    factor(some_vector,
           ordered = TRUE,
           levels = c("lev1", "lev2" ...))

By setting the argument `ordered` to `TRUE` in the function
[`factor()`](http://www.rdocumentation.org/packages/base/functions/factor),
you indicate that the factor is ordered. With the argument `levels` you
give the values of the factor in the correct order.

From `speed_vector`, create an ordered factor vector:
`factor_speed_vector`. Set `ordered` to `TRUE`, and set `levels` to
`c("slow", "medium", "fast")`.

```{r, eval=F}
# Create speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")

# Convert speed_vector to ordered factor vector
factor_speed_vector <- factor(speed_vector, ordered = TRUE, levels = c("slow", "medium", "fast"))

# Print factor_speed_vector
factor_speed_vector
summary(factor_speed_vector)
```

## Comparing ordered factors

Having a bad day at work, 'data analyst number two' enters your office
and starts complaining that 'data analyst number five' is slowing down
the entire project. Since you know that 'data analyst number two' has
the reputation of being a smarty-pants, you first decide to check if his
statement is true.

The fact that `factor_speed_vector` is now ordered enables us to compare
different elements (the data analysts in this case). You can simply do
this by using the well-known operators.

-   Use `[2]` to select from `factor_speed_vector` the factor value for
    the second data analyst. Store it as `da2`.
-   Use `[5]` to select the `factor_speed_vector` factor value for the
    fifth data analyst. Store it as `da5`.
-   Check if `da2` is greater than `da5`; simply print out the result.
    Remember that you can use the `>` operator to check whether one
    element is larger than the other.

```{r, eval=F}
# Create factor_speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")
factor_speed_vector <- factor(speed_vector, ordered = TRUE, levels = c("slow", "medium", "fast"))

# Factor value for second data analyst
da2 <- factor_speed_vector[2]

# Factor value for fifth data analyst
da5 <- factor_speed_vector[5]

# Is data analyst 2 faster data analyst 5?
da2 > da5
```

# Data frames

Most datasets you will be working with will be stored as data frames. By the end of this chapter, you will be able to create a data frame, select interesting parts of a data frame, and order a data frame according to certain variables.

## What's a data frame?

You may remember from the chapter about matrices that all the elements
that you put in a matrix should be of the same type. Back then, your
dataset on Star Wars only contained numeric elements.

When doing a market research survey, however, you often have questions
such as:

-   'Are you married?' or 'yes/no' questions (`logical`)
-   'How old are you?' (`numeric`)
-   'What is your opinion on this product?' or other 'open-ended'
    questions (`character`)
-   …

The output, namely the respondents' answers to the questions formulated
above, is a dataset of different data types. You will often find
yourself working with datasets that contain different data types instead
of only one.

A data frame has the variables of a dataset as columns and the
observations as rows. This will be a familiar concept for those coming
from different statistical software packages such as SAS or SPSS.

Submit the answer. The data from the built-in example data frame
[`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars)
will be printed to the console.

```{r, eval=F}
# Print out built-in R data frame
mtcars 
```

## Quick, have a look at your dataset

Wow, that is a lot of cars!

Working with large datasets is not uncommon in data analysis. When you
work with (extremely) large datasets and data frames, your first task as
a data analyst is to develop a clear understanding of its structure and
main elements. Therefore, it is often useful to show only a small part
of the entire dataset.

So how to do this in R? Well, the function
[`head()`](http://www.rdocumentation.org/packages/utils/functions/head)
enables you to show the first observations of a data frame. Similarly,
the function
[`tail()`](http://www.rdocumentation.org/packages/utils/functions/head)
prints out the last observations in your dataset.

Both
[`head()`](http://www.rdocumentation.org/packages/utils/functions/head)
and
[`tail()`](http://www.rdocumentation.org/packages/utils/functions/head)
print a top line called the 'header', which contains the names of the
different variables in your dataset.

Call
[`head()`](http://www.rdocumentation.org/packages/utils/functions/head)
on the
[`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars)
dataset to have a look at the header and the first observations.

```{r, eval=F}
# Call head() on mtcars
head(mtcars)
```

## Have a look at the structure

Another method that is often used to get a rapid overview of your data
is the function
[`str()`](http://www.rdocumentation.org/packages/utils/functions/str).
The function
[`str()`](http://www.rdocumentation.org/packages/utils/functions/str)
shows you the structure of your dataset. For a data frame it tells you:

-   The total number of observations (e.g. 32 car types)
-   The total number of variables (e.g. 11 car features)
-   A full list of the variables names (e.g. `mpg`, `cyl` … )
-   The data type of each variable (e.g. `num`)
-   The first observations

Applying the
[`str()`](http://www.rdocumentation.org/packages/utils/functions/str)
function will often be the first thing that you do when receiving a new
dataset or data frame. It is a great way to get more insight in your
dataset before diving into the real analysis.

Investigate the structure of
[`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars).
Make sure that you see the same numbers, variables and data types as
mentioned above.

```{r, eval=F}
# Investigate the structure of mtcars
str(mtcars)
```

## Creating a data frame

Since using built-in datasets is not even half the fun of creating your
own datasets, the rest of this chapter is based on your personally
developed dataset. Put your jet pack on because it is time for some
space exploration!

As a first goal, you want to construct a data frame that describes the
main characteristics of eight planets in our solar system. According to
your good friend Buzz, the main features of a planet are:

-   The type of planet (Terrestrial or Gas Giant).
-   The planet's diameter relative to the diameter of the Earth.
-   The planet's rotation across the sun relative to that of the Earth.
-   If the planet has rings or not (TRUE or FALSE).

After doing some high-quality research on
[Wikipedia](https://en.wikipedia.org/wiki/Planet), you feel confident
enough to create the necessary vectors: `name`, `type`, `diameter`,
`rotation` and `rings`; these vectors have already been coded up in the
editor. The first element in each of these vectors correspond to the
first observation.

You construct a data frame with the
[`data.frame()`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/data.frame)
function. As arguments, you pass the vectors from before: they will
become the different columns of your data frame. Because every column
has the same length, the vectors you pass should also have the same
length. But don't forget that it is possible (and likely) that they
contain different types of data.

Use the function
[`data.frame()`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/data.frame)
to construct a data frame. Pass the vectors `name`, `type`, `diameter`,
`rotation` and `rings` as arguments to `data.frame()`, in this order.
Call the resulting data frame `planets_df`.

```{r, eval=F}
# Definition of vectors
name <- c("Mercury", "Venus", "Earth", 
          "Mars", "Jupiter", "Saturn", 
          "Uranus", "Neptune")
type <- c("Terrestrial planet", 
          "Terrestrial planet", 
          "Terrestrial planet", 
          "Terrestrial planet", "Gas giant", 
          "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 
              11.209, 9.449, 4.007, 3.883)
rotation <- c(58.64, -243.02, 1, 1.03,
              0.41, 0.43, -0.72, 0.67)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

# Create a data frame from the vectors
planets_df <- data.frame(name, type, diameter, rotation, rings)
```

## Creating a data frame (2)

The `planets_df` data frame should have 8 observations and 5 variables.
It has been made available in the workspace, so you can directly use it.

Use
[`str()`](http://www.rdocumentation.org/packages/utils/functions/str) to
investigate the structure of the new `planets_df` variable.

```{r, eval=F}
# Check the structure of planets_df
str(planets_df)
```

## Selection of data frame elements

Similar to vectors and matrices, you select elements from a data frame
with the help of square brackets `[ ]`. By using a comma, you can
indicate what to select from the rows and the columns respectively. For
example:

-   `my_df[1,2]` selects the value at the first row and second column in
    `my_df`.
-   `my_df[1:3,2:4]` selects rows 1, 2, 3 and columns 2, 3, 4 in
    `my_df`.

Sometimes you want to select all elements of a row or column. For
example, `my_df[1, ]` selects all elements of the first row. Let us now
apply this technique on `planets_df`!

-   From `planets_df`, select the diameter of Mercury: this is the value
    at the first row and the third column. Simply print out the result.
-   From `planets_df`, select all data on Mars (the fourth row). Simply
    print out the result.

```{r, eval=F}
# The planets_df data frame from the previous exercise is pre-loaded

# Print out diameter of Mercury (row 1, column 3)
planets_df[1,3]

# Print out data for Mars (entire fourth row)
planets_df[4, ]
```

## Selection of data frame elements (2)

Instead of using numerics to select elements of a data frame, you can
also use the variable names to select columns of a data frame.

Suppose you want to select the first three elements of the `type`
column. One way to do this is

    planets_df[1:3,2]

A possible disadvantage of this approach is that you have to know (or
look up) the column number of `type`, which gets hard if you have a lot
of variables. It is often easier to just make use of the variable name:

    planets_df[1:3,"type"]

Select and print out the first 5 values in the `"diameter"` column of
`planets_df`.

```{r, eval=F}
# The planets_df data frame from the previous exercise is pre-loaded

# Select first 5 values of diameter column
planets_df[1:5, "diameter"]
```

## Only planets with rings

You will often want to select an entire column, namely one specific
variable from a data frame. If you want to select all elements of the
variable `diameter`, for example, both of these will do the trick:

    planets_df[,3]
    planets_df[,"diameter"]

However, there is a short-cut. If your columns have names, you can use
the `$` sign:

    planets_df$diameter

-   Use the `$` sign to select the `rings` variable from `planets_df`.
    Store the vector that results as `rings_vector`.
-   Print out `rings_vector` to see if you got it right.

```{r, eval=F}
# planets_df is pre-loaded in your workspace

# Select the rings variable from planets_df
rings_vector <- planets_df$rings

# Print out rings_vector
rings_vector
```

## Only planets with rings (2)

You probably remember from high school that some planets in our solar
system have rings and others do not. Unfortunately you can not recall
their names. Could R help you out?

If you type `rings_vector` in the console, you get:

    [1] FALSE FALSE FALSE FALSE  TRUE  TRUE  TRUE  TRUE

This means that the first four observations (or planets) do not have a
ring (`FALSE`), but the other four do (`TRUE`). However, you do not get
a nice overview of the names of these planets, their diameter, etc.
Let's try to use `rings_vector` to select the data for the four planets
with rings.

The code in the editor selects the `name` column of all planets that
have rings. Adapt the code so that instead of only the `name` column,
*all* columns for planets that have rings are selected.

```{r, eval=F}
# planets_df and rings_vector are pre-loaded in your workspace

# Adapt the code to select all columns for planets with rings
planets_df[rings_vector, ]
```

## Only planets with rings but shorter

So what exactly did you learn in the previous exercises? You selected a
subset from a data frame (`planets_df`) based on whether or not a
certain condition was true (rings or no rings), and you managed to pull
out all relevant data. Pretty awesome! By now, NASA is probably already
flirting with your CV ;-).

Now, let us move up one level and use the function
[`subset()`](http://www.rdocumentation.org/packages/base/functions/subset).
You should see the
[`subset()`](http://www.rdocumentation.org/packages/base/functions/subset)
function as a short-cut to do exactly the same as what you did in the
previous exercises.

    subset(my_df, subset = some_condition)

The first argument of
[`subset()`](http://www.rdocumentation.org/packages/base/functions/subset)
specifies the dataset for which you want a subset. By adding the second
argument, you give R the necessary information and conditions to select
the correct subset.

The code below will give the exact same result as you got in the
previous exercise, but this time, you didn't need the `rings_vector`!

    subset(planets_df, subset = rings)

Use `subset()` on `planets_df` to select planets that have a diameter
smaller than Earth. Because the `diameter` variable is a relative
measure of the planet's diameter w.r.t that of planet Earth, your
condition is `diameter < 1`.

```{r, eval=F}
# planets_df is pre-loaded in your workspace

# Select planets with diameter < 1
subset(planets_df, subset = diameter < 1)
```

## Sorting

Making and creating rankings is one of mankind's favorite affairs. These
rankings can be useful (best universities in the world), entertaining
(most influential movie stars) or pointless (best 007 look-a-like).

In data analysis you can sort your data according to a certain variable
in the dataset. In R, this is done with the help of the function
[`order()`](http://www.rdocumentation.org/packages/base/functions/order).

[`order()`](http://www.rdocumentation.org/packages/base/functions/order)
is a function that gives you the ranked position of each element when it
is applied on a variable, such as a vector for example:

    a <- c(100, 10, 1000)
    order(a)
    [1] 2 1 3

10, which is the second element in `a`, is the smallest element, so 2
comes first in the output of `order(a)`. 100, which is the first element
in `a` is the second smallest element, so 1 comes second in the output
of `order(a)`.

This means we can use the output of `order(a)` to reshuffle `a`:

    a[order(a)]
    [1]   10  100 1000

Experiment with the
[`order()`](http://www.rdocumentation.org/packages/base/functions/order)
function in the console. Submit the answer when you are ready to
continue.

```{r, eval=F}
# Play around with the order function in the console
```

## Sorting your data frame

Alright, now that you understand the
[`order()`](http://www.rdocumentation.org/packages/base/functions/order)
function, let us do something useful with it. You would like to
rearrange your data frame such that it starts with the smallest planet
and ends with the largest one. A sort on the `diameter` column.

-   Call `order()` on `planets_df$diameter` (the `diameter` column of
    `planets_df`). Store the result as `positions`.
-   Now reshuffle `planets_df` with the `positions` vector as row
    indexes inside square brackets. Keep all columns. Simply print out
    the result.

```{r, eval=F}
# planets_df is pre-loaded in your workspace

# Use order() to create positions
positions <- order(planets_df$diameter)

# Use positions to sort planets_df
planets_df[positions, ]
```

# Lists

As opposed to vectors, lists can hold components of different types, just as your to-do lists can contain different categories of tasks. This chapter will teach you how to create, name, and subset these lists.

## Lists, why would you need them?

Congratulations! At this point in the course you are already familiar
with:

-   **Vectors** (one dimensional array): can hold numeric, character or
    logical values. The elements in a vector all have the same data
    type.
-   **Matrices** (two dimensional array): can hold numeric, character or
    logical values. The elements in a matrix all have the same data
    type.
-   **Data frames** (two-dimensional objects): can hold numeric,
    character or logical values. Within a column all elements have the
    same data type, but different columns can be of different data type.

Pretty sweet for an R newbie, right? ;-)

Submit the answer to start learning everything about lists!

```{r, eval=F}
# Just submit the answer.
```

## Lists, why would you need them? (2)

A **list** in R is similar to your to-do list at work or school: the
different items on that list most likely differ in length,
characteristic, and type of activity that has to be done.

A list in R allows you to gather a variety of objects under one name
(that is, the name of the list) in an ordered way. These objects can be
matrices, vectors, data frames, even other lists, etc. It is not even
required that these objects are related to each other in any way.

You could say that a list is some kind super data type: you can store
practically any piece of information in it!

Just submit the answer to start the first exercise on lists.

```{r, eval=F}
# Just submit the answer to start the first exercise on lists.
```

## Creating a list

Let us create our first list! To construct a list you use the function
[`list()`](http://www.rdocumentation.org/packages/base/functions/list):

    my_list <- list(comp1, comp2 ...)

The arguments to the list function are the list components. Remember,
these components can be matrices, vectors, other lists, …

Construct a list, named `my_list`, that contains the variables
`my_vector`, `my_matrix` and `my_df` as list components.

```{r, eval=F}
# Vector with numerics from 1 up to 10
my_vector <- 1:10 

# Matrix with numerics from 1 up to 9
my_matrix <- matrix(1:9, ncol = 3)

# First 10 elements of the built-in data frame mtcars
my_df <- mtcars[1:10,]

# Construct list with these different elements:
my_list <- list(my_vector, my_matrix, my_df)
```

## Creating a named list

Well done, you're on a roll!

Just like on your to-do list, you want to avoid not knowing or
remembering what the components of your list stand for. That is why you
should give names to them:

    my_list <- list(name1 = your_comp1, 
                    name2 = your_comp2)

This creates a list with components that are named `name1`, `name2`, and
so on. If you want to name your lists after you've created them, you can
use the `names()` function as you did with vectors. The following
commands are fully equivalent to the assignment above:

    my_list <- list(your_comp1, your_comp2)
    names(my_list) <- c("name1", "name2")

-   Change the code of the previous exercise (see editor) by adding
    names to the components. Use for `my_vector` the name `vec`, for
    `my_matrix` the name `mat` and for `my_df` the name `df`.
-   Print out `my_list` so you can inspect the output.

```{r, eval=F}
# Vector with numerics from 1 up to 10
my_vector <- 1:10 

# Matrix with numerics from 1 up to 9
my_matrix <- matrix(1:9, ncol = 3)

# First 10 elements of the built-in data frame mtcars
my_df <- mtcars[1:10,]

# Adapt list() call to give the components names
my_list <- list(vec = my_vector, mat = my_matrix, df = my_df)

# Print out my_list
my_list
```

## Creating a named list (2)

Being a huge movie fan (remember your job at LucasFilms), you decide to
start storing information on good movies with the help of lists.

Start by creating a list for the movie "The Shining". We have already
created the variables `mov`, `act` and `rev` in your R workspace. Feel
free to check them out in the console.

Complete the code in the editor to create `shining_list`; it contains
three elements:

-   `moviename`: a character string with the movie title (stored in
    `mov`)
-   `actors`: a vector with the main actors' names (stored in `act`)
-   `reviews`: a data frame that contains some reviews (stored in `rev`)

Do not forget to name the list components accordingly (names are
`moviename`, `actors` and `reviews`).

```{r, eval=F}
# edited/added
mov="The Shining"
act=c("Jack Nicholson","Shelley Duvall","Danny Lloyd","Scatman Crothers","Barry Nelson")
rev=data.frame(scores=c(4.5,4,5),sources=c("IMDb1","IMDb2","IMDb3"),comments=c("Best Horror Film I Have Ever Seen","A truly brilliant and scary film from Stanley Kubrick","A masterpiece of psychological horror"))

# The variables mov, act and rev are available

# Finish the code to build shining_list
shining_list <- list(moviename = mov, actors = act, reviews = rev)
```

## Selecting elements from a list

Your list will often be built out of numerous elements and components.
Therefore, getting a single element, multiple elements, or a component
out of it is not always straightforward.

One way to select a component is using the numbered position of that
component. For example, to "grab" the first component of `shining_list`
you type

    shining_list[[1]]

A quick way to check this out is typing it in the console. Important to
remember: to select elements from vectors, you use single square
brackets: `[ ]`. Don't mix them up!

You can also refer to the names of the components, with `[[ ]]` or with
the `$` sign. Both will select the data frame representing the reviews:

    shining_list[["reviews"]]
    shining_list$reviews

Besides selecting components, you often need to select specific elements
out of these components. For example, with `shining_list[[2]][1]` you
select from the second component, `actors` (`shining_list[[2]]`), the
first element (`[1]`). When you type this in the console, you will see
the answer is Jack Nicholson.

-   Select from `shining_list` the vector representing the actors.
    Simply print out this vector.
-   Select from `shining_list` the second element in the vector
    representing the actors. Do a printout like before.

```{r, eval=F}
# shining_list is already pre-loaded in the workspace

# Print out the vector representing the actors
shining_list$actors

# Print the second element of the vector representing the actors
shining_list$actors[2]
```

## Creating a new list for another movie

You found reviews of another, more recent, Jack Nicholson movie: The
Departed!

| Scores | Comments               |
|--------|------------------------|
| 4.6    | I would watch it again |
| 5      | Amazing!               |
| 4.8    | I liked it             |
| 5      | One of the best movies |
| 4.2    | Fascinating plot       |

It would be useful to collect together all the pieces of information
about the movie, like the title, actors, and reviews into a single
variable. Since these pieces of data are different shapes, it is natural
to combine them in a list variable.

`movie_title`, containing the title of the movie, and `movie_actors`,
containing the names of some of the actors in the movie, are available
in your workspace.

-   Create two vectors, called `scores` and `comments`, that contain the
    information from the reviews shown in the table.
-   Find the average of the `scores` vector and save it as `avg_review`.
-   Combine the `scores` and `comments` vectors into a data frame called
    `reviews_df`.
-   Create a list, called `departed_list`, that contains the
    `movie_title`, `movie_actors`, reviews data frame as `reviews_df`,
    and the average review score as `avg_review`, and print it out.

```{r, eval=F}
# edited/added
movie_title="The Departed"
movie_actors= c("Leonardo DiCaprio","Matt Damon","Jack Nicholson","Mark Wahlberg","Vera Farmiga","Martin Sheen")

# Use the table from the exercise to define the comments and scores vectors
scores <- c(4.6, 5, 4.8, 5, 4.2)
comments <- c("I would watch it again", "Amazing!", "I liked it", "One of the best movies", "Fascinating plot")

# Save the average of the scores vector as avg_review
avg_review <- mean(scores)

# Combine scores and comments into the reviews_df data frame
reviews_df <- data.frame(scores, comments)

# Create and print out a list, called departed_list
departed_list <- list(movie_title, movie_actors, reviews_df, avg_review)
departed_list
```