**Course Description**

This is an introduction to the programming language R, focused on a powerful set of tools known as the Tidyverse. You'll learn the intertwined processes of data manipulation and visualization using the tools dplyr and ggplot2. You'll learn to manipulate data by filtering, sorting, and summarizing a real dataset of historical country data in order to answer exploratory questions. You'll then learn to turn this processed data into informative line plots, bar plots, histograms, and more with the ggplot2 package. You’ll get a taste of the value of exploratory data analysis and the power of Tidyverse tools. This is a suitable introduction for those who have no previous experience in R and are interested in performing data analysis.

# Data wrangling

In this chapter, you'll learn to do three things with a table: filter for particular observations, arrange the observations in a desired order, and mutate to add or change a column. You'll see how each of these steps allows you to answer questions about your data.

## The gapminder dataset

### Loading the gapminder and dplyr packages

Before you can work with the `gapminder` dataset, you'll need to load
two R packages that contain the tools for working with it, then display
the `gapminder` dataset so that you can see what it contains.

To your right, you'll see two windows inside which you can enter code:
The `script.R` window, and the R Console. All of your code to solve each
exercise must go inside `script.R`.

If you hit *Submit Answer*, your R script is executed and the output is
shown in the R Console. DataCamp checks whether your submission is
correct and gives you feedback. You can hit *Submit Answer* as often as
you want. If you're stuck, you can ask for a hint or a solution.

You can use the R Console interactively by simply typing R code and
hitting Enter. When you work in the console directly, your code will not
be checked for correctness so it is a great way to experiment and
explore.

*This course introduces a lot of new concepts, so if you ever need a
quick refresher, download the [Tidyverse For Beginners Cheat
Sheet](https://res.cloudinary.com/dyd911kmh/image/upload/v1676302697/Marketing/Blog/Tidyverse_Cheat_Sheet.pdf)
and keep it handy!*

-   Use the `library()` function to load the `dplyr` package, just like
    we've loaded the `gapminder` package for you.
-   Type `gapminder`, on its own line, to look at the gapminder dataset.

```{r}
# Load the gapminder package
library(gapminder)

# Load the dplyr package
library(dplyr)

# Look at the gapminder dataset
gapminder
```

### Understanding a data frame

Now that you've loaded the `gapminder` dataset, you can start examining
and understanding it.

We've already loaded the `gapminder` and `dplyr` packages. Type
`gapminder` in the console, to display the object.

How many observations (rows) are in the dataset?

- [x] 1704
- [ ] 6
- [ ] 1694
- [ ] 1952

## The filter verb

### Filtering for one year

The `filter` verb extracts particular observations based on a condition.
In this exercise you'll filter for observations from a particular year.

-   Add a `filter()` line after the pipe (`%>%`) to extract only the
    observations from the year 1957. Remember that you use `==` to
    compare two values.

```{r}
library(gapminder)
library(dplyr)

# Filter the gapminder dataset for the year 1957
gapminder %>%
  filter(year == 1957)
```

### Filtering for one country and one year

You can also use the `filter()` verb to set two conditions, which could
retrieve a single observation.

Just like in the last exercise, you can do this in two lines of code,
starting with `gapminder %>%` and having the `filter()` on the second
line. Keeping one verb on each line helps keep the code readable. Note
that each time, you'll put the pipe `%>%` at the end of the first line
(like `gapminder %>%`); putting the pipe at the beginning of the second
line will throw an error.

-   Filter the `gapminder` data to retrieve only the observation from
    China in the year 2002.

```{r}
library(gapminder)
library(dplyr)

# Filter for China in 2002
gapminder %>%
  filter(country == "China", year == 2002)
```

## The arrange verb

### Arranging observations by life expectancy

You use `arrange()` to sort observations in ascending or descending
order of a particular variable. In this case, you'll sort the dataset
based on the `lifeExp` variable.

-   Sort the `gapminder` dataset in ascending order of life expectancy
    (`lifeExp`).
-   Sort the `gapminder` dataset in descending order of life expectancy.

```{r}
library(gapminder)
library(dplyr)

# Sort in ascending order of lifeExp
gapminder %>%
  arrange(lifeExp)
  
# Sort in descending order of lifeExp
gapminder %>%
  arrange(desc(lifeExp))
```

### Filtering and arranging

You'll often need to use the pipe operator (`%>%`) to combine multiple
dplyr verbs in a row. In this case, you'll combine a `filter()` with an
`arrange()` to find the highest population countries in a particular
year.

-   Use `filter()` to extract observations from just the year 1957, then
    use `arrange()` to sort in descending order of population (`pop`).

```{r}
library(gapminder)
library(dplyr)

# Filter for the year 1957, then arrange in descending order of population
gapminder %>%
  filter(year == 1957) %>%
  arrange(desc(pop))
```

## The mutate verb

### Using mutate to change or create a column

Suppose we want life expectancy to be measured in months instead of
years: you'd have to multiply the existing value by 12. You can use the
`mutate()` verb to change this column, or to create a new column that's
calculated this way.

-   Use `mutate()` to change the existing `lifeExp` column, by
    multiplying it by 12: `12 * lifeExp`.
-   Use `mutate()` to add a new column, called `lifeExpMonths`,
    calculated as `12 * lifeExp`.

```{r}
library(gapminder)
library(dplyr)

# Use mutate to change lifeExp to be in months
gapminder %>%
  mutate(lifeExp = lifeExp * 12)

# Use mutate to create a new column called lifeExpMonths
gapminder %>%
  mutate(lifeExpMonths = lifeExp * 12)
```

### Combining filter, mutate, and arrange

In this exercise, you'll combine all three of the verbs you've learned
in this chapter, to find the countries with the highest life expectancy,
in months, in the year 2007.

-   In one sequence of pipes on the `gapminder` dataset:
-   `filter()` for observations from the year 2007,
-   `mutate()` to create a column `lifeExpMonths`, calculated as
    `12 * lifeExp`, and
-   `arrange()` in descending order of that new column

```{r}
library(gapminder)
library(dplyr)

# Filter, mutate, and arrange the gapminder dataset
gapminder %>%
  filter(year == 2007) %>%
  mutate(lifeExpMonths = 12 * lifeExp) %>%
  arrange(desc(lifeExpMonths))
```

# Data visualization

Often a better way to understand and present data as a graph. In this chapter, you'll learn the essential skills of data visualization using the ggplot2 package, and you'll see how the dplyr and ggplot2 packages work closely together to create informative graphs.

## Visualizing with ggplot2

### Variable assignment

Throughout the exercises in this chapter, you'll be visualizing a subset
of the gapminder data from the year 1952. First, you'll have to load the
ggplot2 package, and create a `gapminder_1952` dataset to visualize.

*By the way, if you haven't downloaded it already, check out the
[Tidyverse For Beginners Cheat
Sheet](https://res.cloudinary.com/dyd911kmh/image/upload/v1676302697/Marketing/Blog/Tidyverse_Cheat_Sheet.pdf).
It includes an overview of the most important concepts, functions and
methods and might come in handy if you ever need a quick refresher!*

-   Load the `ggplot2` package after the gapminder and dplyr packages.
-   Filter `gapminder` for observations from the year 1952, and assign
    it to a new dataset `gapminder_1952` using the assignment operator
    (`<-`).

```{r}
# Load the ggplot2 package as well
library(gapminder)
library(dplyr)
library(ggplot2)

# Create gapminder_1952
gapminder_1952 <- gapminder %>%
  filter(year == 1952)
```

### Comparing population and GDP per capita

In the video you learned to create a scatter plot with GDP per capita on
the x-axis and life expectancy on the y-axis (the code for that graph
has been provided in the exercise code). When you're exploring data
visually, you'll often need to try different combinations of variables
and aesthetics.

-   Change the scatter plot of `gapminder_1952` so that (`pop`) is on
    the x-axis and GDP per capita (`gdpPercap`) is on the y-axis.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Change to put pop on the x-axis and gdpPercap on the y-axis
ggplot(gapminder_1952, aes(x = pop, y = gdpPercap)) +
  geom_point()
```

### Comparing population and life expectancy

In this exercise, you'll use `ggplot2` to create a scatter plot from
scratch, to compare each country's population with its life expectancy
in the year 1952.

-   Create a scatter plot of `gapminder_1952` with population (`pop`) is
    on the x-axis and life expectancy (`lifeExp`) on the y-axis.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Create a scatter plot with pop on the x-axis and lifeExp on the y-axis
ggplot(gapminder_1952, aes(x = pop, y = lifeExp)) +
  geom_point()
```

## Log scales

### Putting the x-axis on a log scale

You previously created a scatter plot with population on the x-axis and
life expectancy on the y-axis. Since population is spread over several
orders of magnitude, with some countries having a much higher population
than others, it's a good idea to put the x-axis on a log scale.

-   Change the existing scatter plot (code provided) to put the x-axis
    (representing population) on a log scale.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Change this plot to put the x-axis on a log scale
ggplot(gapminder_1952, aes(x = pop, y = lifeExp)) +
  geom_point() +
  scale_x_log10()
```

### Putting the x- and y- axes on a log scale

Suppose you want to create a scatter plot with population on the x-axis
and GDP per capita on the y-axis. Both population and GDP per-capita are
better represented with log scales, since they vary over many orders of
magnitude.

-   Create a scatter plot with population (`pop`) on the x-axis and GDP
    per capita (`gdpPercap`) on the y-axis. Put **both** the x- and y-
    axes on a log scale.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Scatter plot comparing pop and gdpPercap, with both axes on a log scale
ggplot(gapminder_1952, aes(x = pop, y = gdpPercap)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

## Additional aesthetics

### Adding color to a scatter plot

In this lesson you learned how to use the color aesthetic, which can be
used to show which continent each point in a scatter plot represents.

-   Create a scatter plot with population (`pop`) on the x-axis, life
    expectancy (`lifeExp`) on the y-axis, and with continent
    (`continent`) represented by the color of the points. Put the x-axis
    on a log scale.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Scatter plot comparing pop and lifeExp, with color representing continent
ggplot(gapminder_1952, aes(x = pop, y = lifeExp, color = continent)) +
  geom_point() +
  scale_x_log10()
```

### Adding size and color to a plot

In the last exercise, you created a scatter plot communicating
information about each country's population, life expectancy, and
continent. Now you'll use the size of the points to communicate even
more.

-   Modify the scatter plot so that the size of the points represents
    each country's GDP per capita (`gdpPercap`).

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Add the size aesthetic to represent a country's gdpPercap
ggplot(gapminder_1952, aes(x = pop, y = lifeExp, color = continent, size = gdpPercap)) +
  geom_point() +
  scale_x_log10()
```

## Faceting

### Creating a subgraph for each continent

You've learned to use faceting to divide a graph into subplots based on
one of its variables, such as the continent.

-   Create a scatter plot of `gapminder_1952` with the x-axis
    representing population (`pop`), the y-axis representing life
    expectancy (`lifeExp`), and faceted to have one subplot per
    continent (`continent`). Put the x-axis on a log scale.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Scatter plot comparing pop and lifeExp, faceted by continent
ggplot(gapminder_1952, aes(x = pop, y = lifeExp)) +
  geom_point() +
  scale_x_log10() +
  facet_wrap(~ continent)
```

### Faceting by year

All of the graphs in this chapter have been visualizing statistics
within one year. Now that you're able to use faceting, however, you can
create a graph showing **all** the country-level data from 1952 to 2007,
to understand how global statistics have changed over time.

-   Create a scatter plot of the `gapminder` data:
-   Put GDP per capita (`gdpPercap`) on the x-axis and life expectancy
    (`lifeExp`) on the y-axis, with continent (`continent`) represented
    by color and population (`pop`) represented by size.
-   Put the x-axis on a log scale
-   Facet by the `year` variable

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

# Scatter plot comparing gdpPercap and lifeExp, with color representing continent
# and size representing population, faceted by year
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
  geom_point() +
  scale_x_log10() +
  facet_wrap(~ year)
```

# Grouping and summarizing

So far you've been answering questions about individual country-year pairs, but you may be interested in aggregations of the data, such as the average life expectancy of all countries within each year. Here you'll learn to use the group by and summarize verbs, which collapse large datasets into manageable summaries.

## The summarize verb

### Summarizing the median life expectancy

You've seen how to find the mean life expectancy and the total
population across a set of observations, but `mean()` and `sum()` are
only two of the functions R provides for summarizing a collection of
numbers. Here, you'll learn to use the `median()` function in
combination with `summarize()`.

By the way, `dplyr` displays some messages when it's loaded that we've
been hiding so far. They'll show up in red and start with:

    Attaching package: 'dplyr'

    The following objects are masked from 'package:stats':

This will occur in future exercises each time you load `dplyr`: it's
mentioning some built-in functions that are overwritten by `dplyr`. You
won't need to worry about this message within this course.

-   Use the `median()` function within a `summarize()` to find the
    median life expectancy. Save it into a column called
    `medianLifeExp`.

```{r}
library(gapminder)
library(dplyr)

# Summarize to find the median life expectancy
gapminder %>%
  summarize(medianLifeExp = median(lifeExp))
```

### Summarizing the median life expectancy in 1957

Rather than summarizing the entire dataset, you may want to find the
median life expectancy for only one particular year. In this case,
you'll find the median in the year 1957.

-   Filter for the year 1957, then use the `median()` function within a
    `summarize()` to calculate the median life expectancy into a column
    called `medianLifeExp`.

```{r}
library(gapminder)
library(dplyr)

# Filter for 1957 then summarize the median life expectancy
gapminder %>%
  filter(year == 1957) %>%
  summarize(medianLifeExp = median(lifeExp))
```

### Summarizing multiple variables in 1957

The `summarize()` verb allows you to summarize multiple variables at
once. In this case, you'll use the `median()` function to find the
median life expectancy and the `max()` function to find the maximum GDP
per capita.

-   Find both the median life expectancy (`lifeExp`) and the maximum GDP
    per capita (`gdpPercap`) in the year 1957, calling them
    `medianLifeExp` and `maxGdpPercap` respectively. You can use the
    `max()` function to find the maximum.

```{r}
library(gapminder)
library(dplyr)

# Filter for 1957 then summarize the median life expectancy and the maximum GDP per capita
gapminder %>%
  filter(year == 1957) %>%
  summarize(medianLifeExp = median(lifeExp),
            maxGdpPercap = max(gdpPercap))
```

## The group_by verb

### Summarizing by year

In a previous exercise, you found the median life expectancy and the
maximum GDP per capita in the year 1957. Now, you'll perform those two
summaries within each year in the dataset, using the `group_by` verb.

-   Find the median life expectancy (`lifeExp`) and maximum GDP per
    capita (`gdpPercap`) **within each year**, saving them into
    `medianLifeExp` and `maxGdpPercap`, respectively.

```{r}
library(gapminder)
library(dplyr)

# Find median life expectancy and maximum GDP per capita in each year
gapminder %>%
  group_by(year) %>%
  summarize(medianLifeExp = median(lifeExp),
            maxGdpPercap = max(gdpPercap))
```

### Summarizing by continent

You can group by any variable in your dataset to create a summary.
Rather than comparing across time, you might be interested in comparing
among continents. You'll want to do that within one year of the dataset:
let's use 1957.

-   Filter the `gapminder` data for the year 1957. Then find the median
    life expectancy (`lifeExp`) and maximum GDP per capita (`gdpPercap`)
    **within each continent**, saving them into `medianLifeExp` and
    `maxGdpPercap`, respectively.

```{r}
library(gapminder)
library(dplyr)

# Find median life expectancy and maximum GDP per capita in each continent in 1957
gapminder %>%
  filter(year == 1957) %>%
  group_by(continent) %>%
  summarize(medianLifeExp = median(lifeExp),
            maxGdpPercap = max(gdpPercap))
```

### Summarizing by continent and year

Instead of grouping just by year, or just by continent, you'll now group
by both continent and year to summarize within each.

-   Find the median life expectancy (`lifeExp`) and maximum GDP per
    capita (`gdpPercap`) **within each combination of continent and
    year**, saving them into `medianLifeExp` and `maxGdpPercap`,
    respectively.

```{r}
library(gapminder)
library(dplyr)

# Find median life expectancy and maximum GDP per capita in each continent/year combination
gapminder %>%
  group_by(continent, year) %>%
  summarize(medianLifeExp = median(lifeExp),
            maxGdpPercap = max(gdpPercap))
```

## Visualizing summarized data

### Visualizing median life expectancy over time

In the last chapter, you summarized the gapminder data to calculate the
median life expectancy within each year. This code is provided for you,
and is saved (with `<-`) as the `by_year` dataset.

Now you can use the ggplot2 package to turn this into a visualization of
changing life expectancy over time.

-   Use the `by_year` dataset to create a scatter plot showing the
    change of median life expectancy over time, with `year` on the
    x-axis and `medianLifeExp` on the y-axis. Be sure to add
    `expand_limits(y = 0)` to make sure the plot's y-axis includes zero.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

by_year <- gapminder %>%
  group_by(year) %>%
  summarize(medianLifeExp = median(lifeExp),
            maxGdpPercap = max(gdpPercap))

# Create a scatter plot showing the change in medianLifeExp over time
ggplot(by_year, aes(x = year, y = medianLifeExp)) +
  geom_point() +
  expand_limits(y = 0)
```

### Visualizing median GDP per capita per continent over time

In the last exercise you were able to see how the median life expectancy
of countries changed over time. Now you'll examine the median GDP per
capita instead, and see how the trend differs among continents.

-   Summarize the gapminder dataset by continent and year, finding the
    median GDP per capita (`gdpPercap`) within each and putting it into
    a column called `medianGdpPercap`. Use the assignment operator `<-`
    to save this summarized data as `by_year_continent`.
-   Create a scatter plot showing the change in `medianGdpPercap` by
    continent over time. Use color to distinguish between continents,
    and be sure to add `expand_limits(y = 0)` so that the y-axis starts
    at zero.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

# Summarize medianGdpPercap within each continent within each year: by_year_continent
by_year_continent <- gapminder %>%
  group_by(continent, year) %>%
  summarize(medianGdpPercap = median(gdpPercap))

# Plot the change in medianGdpPercap in each continent over time
ggplot(by_year_continent, aes(x = year, y = medianGdpPercap, color = continent)) +
  geom_point() +
  expand_limits(y = 0)
```

### Comparing median life expectancy and median GDP per continent in 2007

In these exercises you've generally created plots that show change over
time. But as another way of exploring your data visually, you can also
use ggplot2 to plot summarized data to compare continents within a
single year.

-   Filter the gapminder dataset for the year 2007, then summarize the
    median GDP per capita and the median life expectancy **within each
    continent**, into columns called `medianLifeExp` and
    `medianGdpPercap`. Save this as `by_continent_2007`.
-   Use the `by_continent_2007` data to create a scatterplot comparing
    these summary statistics for continents in 2007, putting the median
    GDP per capita on the x-axis to the median life expectancy on the
    y-axis. Color the scatter plot by `continent`. You don't need to add
    `expand_limits(y = 0)` for this plot.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

# Summarize the median GDP and median life expectancy per continent in 2007
by_continent_2007 <- gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarize(medianGdpPercap = median(gdpPercap),
            medianLifeExp = median(lifeExp))

# Use a scatter plot to compare the median GDP and median life expectancy
ggplot(by_continent_2007, aes(x = medianGdpPercap, y = medianLifeExp, color = continent)) +
  geom_point()
```

# Types of visualizations

In this chapter, you'll learn how to create line plots, bar plots, histograms, and boxplots. You'll see how each plot requires different methods of data manipulation and preparation, and you’ll understand how each of these plot types plays a different role in data analysis.

## Line plots

### Visualizing median GDP per capita over time

A line plot is useful for visualizing trends over time. In this
exercise, you'll examine how the median GDP per capita has changed over
time.

-   Use `group_by()` and `summarize()` to find the median GDP per capita
    **within each year**, calling the output column `medianGdpPercap`.
    Use the assignment operator `<-` to save it to a dataset called
    `by_year`.
-   Use the `by_year` dataset to create a line plot showing the change
    in median GDP per capita over time. **Be sure** to use
    `expand_limits(y = 0)` to include 0 on the y-axis.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

# Summarize the median gdpPercap by year, then save it as by_year
by_year <- gapminder %>%
  group_by(year) %>%
  summarize(medianGdpPercap = median(gdpPercap))

# Create a line plot showing the change in medianGdpPercap over time
ggplot(by_year, aes(x = year, y = medianGdpPercap)) +
  geom_line() +
  expand_limits(y = 0)
```

### Visualizing median GDP per capita by continent over time

In the last exercise you used a line plot to visualize the increase in
median GDP per capita over time. Now you'll examine the change within
each continent.

-   Use `group_by()` and `summarize()` to find the median GDP per capita
    **within each year and continent**, calling the output column
    `medianGdpPercap`. Use the assignment operator `<-` to save it to a
    dataset called `by_year_continent`.
-   Use the `by_year_continent` dataset to create a line plot showing
    the change in median GDP per capita over time, with color
    representing continent. **Be sure** to use `expand_limits(y = 0)` to
    include 0 on the y-axis.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

# Summarize the median gdpPercap by year & continent, save as by_year_continent
by_year_continent <- gapminder %>%
  group_by(year, continent) %>%
  summarize(medianGdpPercap = median(gdpPercap))

# Create a line plot showing the change in medianGdpPercap by continent over time
ggplot(by_year_continent, aes(x = year, y = medianGdpPercap, color = continent)) +
  geom_line() +
  expand_limits(y = 0)
```

## Bar plots

### Visualizing median GDP per capita by continent

A bar plot is useful for visualizing summary statistics, such as the
median GDP in each continent.

-   Use `group_by()` and `summarize()` to find the median GDP per capita
    **within each continent** in the year 1952, calling the output
    column `medianGdpPercap`. Use the assignment operator `<-` to save
    it to a dataset called `by_continent`.
-   Use the `by_continent` dataset to create a bar plot showing the
    median GDP per capita in each continent.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

# Summarize the median gdpPercap by continent in 1952
by_continent <- gapminder %>%
  filter(year == 1952) %>%
  group_by(continent) %>%
  summarize(medianGdpPercap = median(gdpPercap))

# Create a bar plot showing medianGdp by continent
ggplot(by_continent, aes(x = continent, y = medianGdpPercap)) +
  geom_col()
```

### Visualizing GDP per capita by country in Oceania

You've created a plot where each bar represents one continent, showing
the median GDP per capita for each. But the x-axis of the bar plot
doesn't have to be the continent: you can instead create a bar plot
where each bar represents a country.

In this exercise, you'll create a bar plot comparing the GDP per capita
between the two countries in the Oceania continent (Australia and New
Zealand).

-   Filter for observations in the **Oceania** continent in the
    year 1952. Save this as `oceania_1952`.
-   Use the `oceania_1952` dataset to create a bar plot, with country on
    the x-axis and `gdpPercap` on the y-axis.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

# Filter for observations in the Oceania continent in 1952
oceania_1952 <- gapminder %>%
  filter(continent == "Oceania", year == 1952)

# Create a bar plot of gdpPercap by country
ggplot(oceania_1952, aes(x = country, y = gdpPercap)) +
  geom_col()
```

## Histograms

### Visualizing population

A histogram is useful for examining the distribution of a numeric
variable. In this exercise, you'll create a histogram showing the
distribution of country populations (by millions) in the year 1952.

Code for generating this dataset, `gapminder_1952`, is provided.

Use the `gapminder_1952` dataset to create a histogram of country
population (`pop_by_mil`) in the year 1952. Inside the histogram geom,
set the number of `bins` to `50`.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_1952 <- gapminder %>%
  filter(year == 1952) %>%
  mutate(pop_by_mil = pop / 1000000)

# Create a histogram of population (pop_by_mil)
ggplot(gapminder_1952, aes(x = pop_by_mil)) +
  geom_histogram(bins = 50)
```

### Visualizing population with x-axis on a log scale

In the last exercise you created a histogram of populations across
countries. You might have noticed that there were several countries with
a much higher population than others, which causes the distribution to
be very skewed, with most of the distribution crammed into a small part
of the graph. (Consider that it's hard to tell the median or the minimum
population from that histogram).

To make the histogram more informative, you can try putting the x-axis
on a log scale.

-   Use the `gapminder_1952` dataset (code is provided) to create a
    histogram of country population (`pop`) in the year 1952, putting
    the x-axis on a log scale with `scale_x_log10()`.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Create a histogram of population (pop), with x on a log scale
ggplot(gapminder_1952, aes(x = pop)) +
  geom_histogram() +
  scale_x_log10()
```

## Boxplots

### Comparing GDP per capita across continents

A boxplot is useful for comparing a distribution of values across
several groups. In this exercise, you'll examine the distribution of GDP
per capita by continent. Since GDP per capita varies across several
orders of magnitude, you'll need to put the y-axis on a log scale.

-   Use the `gapminder_1952` dataset (code is provided) to create a
    boxplot comparing GDP per capita (`gdpPercap`) among continents. Put
    the y-axis on a log scale with `scale_y_log10()`.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Create a boxplot comparing gdpPercap among continents
ggplot(gapminder_1952, aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  scale_y_log10()
```

### Adding a title to your graph

There are many other options for customizing a `ggplot2` graph, which
you can learn about in other DataCamp courses. You can also learn about
them from online resources, which is an important skill to develop.

As the final exercise in this course, you'll practice looking up
`ggplot2` instructions by completing a task we haven't shown you how to
do.

-   Add a title to the graph: **Comparing GDP per capita across
    continents**. Use a search engine, such as Google or Bing, to learn
    how to do so.
-   After this exercise you are almost done with your course. If you
    enjoyed the material, feel free to send Dave a thank you via
    twitter. He'll appreciate it. [Tweet to
    Dave](http://twitter.com/home?status=Thoroughly%20enjoyed%20the%20Introduction%20to%20the%20Tidyverse%20course%20%40DataCamp%20by%20%40drob.%20Great%20instructor!%20https%3A%2F%2Fbit.ly%2F2AmGt3t%0A)

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Add a title to this graph: "Comparing GDP per capita across continents"
ggplot(gapminder_1952, aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  scale_y_log10() +
  ggtitle("Comparing GDP per capita across continents")
```

## Conclusion

### Conclusion
Congratulations on completing this Introduction to R via the Tidyverse. You've been introduced to the principles

### Transforming and visualizing data with R

of transforming and visualizing data with R, and in the process learned some real insights from the Gapminder dataset. This course forms a great foundation for other DataCamp courses where you can continue learning how to use these powerful tools to explore data. You can take

### Next steps: Data visualization

courses about ggplot2 to learn to create more informative and customized data visualizations. You can learn much more about using dplyr to transform your data, such as how to join multiple tables together. To analyze other data that you're interested in, you can take the course on importing and cleaning datasets. And you can practice your data wrangling and visualization skills in my own course "Exploratory Data Analysis with R", which offers a case study of analyzing United Nations voting over time. These are just a few of the many resources you have to continue learning about data science and R.

### Enjoy your data science journey!

I hope you had fun in this course, and continue to enjoy your data science journey!
